Learning an Intervention via Causal Inference

### 2016-01-31 causal-bandit/icml-2015/general.tex
updates

Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{t,i} = j))= & \nonumber \\
 \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
}

Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. 
If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to 
explicitly play each of these arms individually to estimate their rewards.

@@

Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
We would like to estimate
\eq{
\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
}
using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$.

>>>>

Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. 
If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to 
explicitly play each of these arms individually to estimate their rewards.

@@

Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
We would like to estimate
\eq{
\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
}
using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$. Ultimately we will be interested
in the special case that $Q(x) = P(x|do(X_i = j))$ and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
\eqn{
\label{eq:truncated}
Z = Y R \ind{R < B}\,,
}
which trivially satisfies $|Z| \leq B$ surely.
Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
Computing the expectation we see
\eq{
\EE_P[Z] = \mu - \beta\,.
}
where $\beta \geq 0$ is the negative bias and satisfies
\eq{
\beta \leq \Q{R \geq B}\,.
}
Bounding the variance is also straightforward.
\eq{
\Var[Z] 
&\leq \EE[Z^2] 
=\sum_x \P{X = x} Z(x)^2 \\
&\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 \\
&\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
= V\,.
}
The bias can be controlled in terms of $V$ via Markov's inequality.
\eq{
\beta 
\leq \Q{R \geq B} 
\leq \frac{\EE_Q[R]}{B}
= \frac{V}{B}\,.
}
The point is that $B$ can be tuned to trade the bias against the concentration of the estimator,
the latter of which can be controlled via Bernstein's inequality.

Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.

\begin{algorithm}[H]
\caption{CheckValid}\label{alg:alloc}
\begin{algorithmic}
\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
\FOR{$i \in \set{o_1,\ldots,o_N}$ and $j \in \set{0,1}$}
\STATE Compute variances:
\eq{
\forall k \in \calP_i,\, l \in \set{0,1} \quad
V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
}
\STATE Compute variance for learning from observational data:
\eq{
V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
}
\STATE Compute required budget: 
\eq{
N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
} 
\ENDFOR
\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
\STATE {\bf return true}
\ELSE
\STATE {\bf return false}
\ENDIF
\end{algorithmic}
\end{algorithm}
